# Results
```{r}
# required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggalluvial)
library(patchwork)
library(vcd)
library(forcats)
library(RColorBrewer)
```

```{r}
# Read in data
data <- read.csv("NYPD_Arrest_Data.csv", na.strings = c("(null)", "N/A"))
```

```{r}
### Data Preprocessing step
data <- na.omit(data)
data$ARREST_DATE <- as.Date(data$ARREST_DATE, format = "%m/%d/%Y")
data <- data |>
  mutate(ARREST_BORO = case_when(
    ARREST_BORO == "B" ~ "Bronx",
    ARREST_BORO == "S" ~ "Staten Island",
    ARREST_BORO == "K" ~ "Brooklyn",
    ARREST_BORO == "M" ~ "Manhattan",
    ARREST_BORO == "Q" ~ "Queens"
  )) |>
  mutate(LAW_CAT_CD = case_when(
    LAW_CAT_CD == "F" ~ "Felony",
    LAW_CAT_CD == "M" ~ "Misdemeanor",
    LAW_CAT_CD == "V" ~ "Violation",
    .default = NA
  )) |>
  mutate(JURISDICTION_CODE = case_when(
    JURISDICTION_CODE == 0 ~ "Patrol",
    JURISDICTION_CODE == 1 ~ "Transit",
    JURISDICTION_CODE == 2 ~ "Housing",
    .default = "Non NYPD"
  )) |>
  mutate(PERP_SEX = case_when(
    PERP_SEX == "F" ~ "Female",
    PERP_SEX == "M" ~ "Male"
  ))
```

Crime frequency distribution across precincts in NYC: 
```{r}
daily_data <- data |>
  group_by(ARREST_DATE) |>
  summarise(Count = n(), .groups = "drop")

ggplot(daily_data, aes(x = ARREST_DATE, y = Count)) +
  geom_line(size = 0.8, alpha = 0.8) +
  labs(
    title = "Daily NYPD Arrests",
    subtitle = "From January 1st to September 30th",
    x = "Date",
    y = "Number of Arrests",
    color = "Borough"
  ) +
  scale_x_date(date_breaks = "2 week", date_labels = "%b %d") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

2. Overall demographic distribution of arrests:
```{r}
age_count <- data |>
  group_by(AGE_GROUP) |>
  summarize(Count = n())

plot1 <- ggplot(age_count, aes(x = AGE_GROUP, y = Count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  ggtitle("Number of Arrests by Age group") +
  xlab("Perp Age Group") +
  ylab("Number of Arrests")

gender_counts <- data |>
  group_by(PERP_SEX) |>
  summarize(Count = n()) |>
  mutate(Percentage = Count / sum(Count) * 100,
         Label = paste0(PERP_SEX, " ", round(Percentage, 1), "%"),
         Position = cumsum(Count) - Count / 2)

plot2 <- ggplot(gender_counts, aes(x = 2, y = Count, fill = PERP_SEX)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  labs(title = "Perp Gender") +
  theme_void() +
  scale_fill_manual(values = c("steelblue", "lightblue")) +
  geom_text(aes(x = 3, y = Position, label = Label), color = "black", size = 5) + 
  theme(legend.position = "none") + xlim(0.5, 3)

race_count <- data |>
  group_by(PERP_RACE) |>
  summarize(Count = n()) |>
  arrange(desc(Count))

plot3 <- ggplot(race_count, aes(x = reorder(PERP_RACE, -Count), y = Count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  ggtitle("Number of Arrests by PERP_RACE") +
  xlab("PERP_RACE") +
  ylab("Number of Arrests") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

top_15_ofns_desc <- data |>
  group_by(OFNS_DESC) |>
  summarise(Count = n()) |>
  arrange(desc(Count)) |>
  head(15)

plot4 <- ggplot(top_15_ofns_desc, aes(x = reorder(OFNS_DESC, Count), 
                                      y = Count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(title = "Top 15 Arrest Categories",
       x = "Offense Description", y = "Number of Arrests") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  coord_flip()
```

```{r}
layout <- "\nABD\nCCD"

(plot1 + plot2 + plot3) + plot4 + plot_layout(design = layout, widths = c(1, 1))
```

This plot provides a comprehensive overview of demographic patterns and crime-related data. The top-left plot displays the distribution of arrests by age group, revealing that individuals aged `25-44` represent the highest proportion of arrests, followed by the `18-24` and `45-64` age groups. The pie chart at the top-right illustrates the gender distribution of arrests, with males accounting for a dominant 82.1% of all arrests, compared to 17.9% for females. The bottom-left plot focuses on the perpetrator’s race description, showing that `Black` individuals are most frequently arrested, followed by `White Hispanics`, `Black Hispanics`, and other racial categories. Finally, the bar chart on the bottom-right lists the top 15 offense categories, where `Assault 3 & Related Offenses` leads, followed by `Petit Larceny` and `Felony Assault`.

```{r}
alluvia_data <- na.omit(data) |> 
  filter(!(PERP_RACE %in% c("UNKNOWN", "AMERICAN INDIAN/ALASKAN NATIVE"))) |>
  group_by(PERP_SEX, PERP_RACE, AGE_GROUP) |>
  summarise(Count = n(), .groups = "drop") |>
  rename(Sex = PERP_SEX, Race = PERP_RACE, AgeGroup = AGE_GROUP)

ggplot(alluvia_data, aes(axis1 = Sex, axis2 = Race, 
                         axis3 = AgeGroup, y = Count)) +
  geom_alluvium(aes(fill = Race), width = 0.2, alpha = 0.8) +
  geom_stratum(width = 0.2, fill = "lightgrey", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2) +
  scale_x_discrete(limits = c("Sex", "Race", "AgeGroup"), expand = c(0.1, 0.1)) +
  theme_minimal() +
  scale_fill_brewer(type='qual') + 
  labs(title = "Flow of Arrests by Demographics",
       x = "Demographic Categories",
       y = "Number of Arrests", fill = "Race")
```

Among Black females, the majority of arrests are concentrated in the 25-44 age group, with a smaller proportion in the 18-24 and 45-64 age groups. Arrests of Black females under 18 or over 65 are minimal.

```{r}
# data1 <- data |> filter(!is.na(LAW_CAT_CD)) |> 
#   filter(!(PERP_RACE %in% c("UNKNOWN", "AMERICAN INDIAN/ALASKAN NATIVE"))) |>
#   filter(!LAW_CAT_CD %in% c("Violation")) |>
#   mutate(ARREST_BORO = as.factor(ARREST_BORO),
#          LAW_CAT_CD = as.factor(LAW_CAT_CD),
#          PERP_RACE = as.factor(PERP_RACE))

mosaic_data <- data |> filter(!is.na(LAW_CAT_CD)) |> 
  filter(!(PERP_RACE %in% c("UNKNOWN", "AMERICAN INDIAN/ALASKAN NATIVE")))

fill_colors <- RColorBrewer::brewer.pal(3, "Blues")

# Generate the mosaic plot
mosaic(LAW_CAT_CD ~ ARREST_BORO + PERP_RACE, data = mosaic_data,
       labeling_args = list(
         set_varnames = c(
           ARREST_BORO = "Borough", 
           PERP_RACE = "Race", 
           LAW_CAT_CD = "Offense Level"
         )),
       direction = c("v","h","h"),
       highlighting_fill = fill_colors,
       main = "Mosaic Plot: Borough, Race, and Offense Level")
```

