# Results
library(ggplot2)

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(naniar)
library(ggmap)
library(lubridate)
library(sf)
library(dygraphs)
library(xts)
library(ggiraph)
library(sf)

library(ggalluvial)
library(patchwork)
library(vcd)
library(forcats)
library(RColorBrewer)
```

```{r}
# Read in data
data <- read.csv("NYPD_Arrest_Data.csv", na.strings = c("(null)", "N/A"))
```

```{r}
### Data Preprocessing step
data <- na.omit(data)
data$ARREST_DATE <- as.Date(data$ARREST_DATE, format = "%m/%d/%Y")
data <- data |>
  mutate(ARREST_BORO = case_when(
    ARREST_BORO == "B" ~ "Bronx",
    ARREST_BORO == "S" ~ "Staten Island",
    ARREST_BORO == "K" ~ "Brooklyn",
    ARREST_BORO == "M" ~ "Manhattan",
    ARREST_BORO == "Q" ~ "Queens"
    )) |>
  mutate(LAW_CAT_CD = case_when(
    LAW_CAT_CD == "F" ~ "Felony",
    LAW_CAT_CD == "M" ~ "Misdemeanor",
    LAW_CAT_CD == "V" ~ "Violation",
    .default = NA
  )) |>
  mutate(JURISDICTION_CODE = case_when(
    JURISDICTION_CODE == 0 ~ "Patrol",
    JURISDICTION_CODE == 1 ~ "Transit",
    JURISDICTION_CODE == 2 ~ "Housing",
    .default = "Non NYPD"
  )) |>
  mutate(PERP_SEX = case_when(
    PERP_SEX == "F" ~ "Female",
    PERP_SEX == "M" ~ "Male"
  )) |>
  mutate(Weekday = wday(ARREST_DATE, label = TRUE, abbr = TRUE)) 
```

# Geographical Patterns of Crime in NYC
```{r fig.height=10, fig.width=10}
nyc_sf <- read_sf("new-york-city-boroughs.geojson")
data <- data |>
  filter(Longitude != 0 & Latitude != 0)
arrest_sf <- st_as_sf(data, coords = c("Longitude", "Latitude"), crs = 4326)

borough_arrest_count <- arrest_sf |>
  st_drop_geometry() |>
  group_by(ARREST_BORO) |>
  summarise(total_arrests = n())

nyc_sf <- nyc_sf |>
  left_join(borough_arrest_count, by = c("name" = "ARREST_BORO")) |>
  mutate(tooltip = paste(name, "<br>Total Arrests:", total_arrests))

interactive_map <- ggplot() +
  geom_sf_interactive(data = nyc_sf,
                      aes(fill = name, geometry = geometry, tooltip = tooltip),
                      color = "black", size = 0.3, alpha = 0.5) +
  geom_sf(data = arrest_sf, aes(geometry = geometry),
          color = "red", size = 0.05, alpha = 0.4, stroke = 0.3, shape = 1) +
  labs(title = "Arrest Locations in NYC", 
       x = "Longitude", y = "Latitude", fill = "Borough") +
  coord_sf() +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12, face = "italic"),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        plot.margin = margin(1, 1, 1, 1, "cm"))

girafe(ggobj = interactive_map)
```

Brooklyn leads with the highest number of arrests (53,987), showcasing widespread clusters, particularly in its northern and eastern neighborhoods. These areas may experience higher crime rates due to concentrated urban activity and demographic factors.

Manhattan follows with 46,226 arrests, heavily concentrated in its central and southern regions, including areas like Midtown and Downtown. The high urban density, commercial activity, and large influx of daily commuters likely contribute to this concentration. The Bronx, with 43,896 arrests, shows dense clusters in its central and southern areas, reflecting persistent socio-economic challenges and a relatively high population density.

Queens accounts for 42,969 arrests, characterized by a more dispersed pattern with prominent clusters in neighborhoods like Jamaica and Flushing. The suburban nature of Queens contrasts with the dense urban dynamics of Manhattan and the Bronx. Staten Island, with only 8,341 arrests, has the lowest crime figures, reflecting its smaller population and suburban characteristics, with most arrests concentrated in the northern part of the borough.

```{r}
ggplot(data, aes(x = Longitude, y = Latitude)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon") +
  scale_fill_viridis_c() +
  labs(
    title = "Crime Density Heatmap in NYC",
    x = "Longitude",
    y = "Latitude",
    fill = "Density"
  ) +
  theme_minimal()
```
```{r}
nyc_precincts <- st_read("new-york-city-boroughs.geojson")

# Aggregate arrests by precinct
precinct_arrests <- data |>
  group_by(ARREST_PRECINCT) |>
  summarise(Count = n())

# Merge arrest data with precinct shapefile
precinct_arrests_map <- nyc_precincts |>
  left_join(precinct_arrests, by = c("PRECINCT" = "ARREST_PRECINCT"))

# Plot precinct-level arrests
ggplot(data = precinct_arrests_map) +
  geom_sf(aes(fill = Count), color = "white") +
  scale_fill_viridis_c() +
  labs(
    title = "Arrests by Police Precinct in NYC",
    fill = "Number of Arrests"
  ) +
  theme_minimal()
```

# Demographic Insights into Arrest Trends
```{r}
daily_data <- data |>
  group_by(ARREST_DATE) |>
  summarise(Count = n(), .groups = "drop")
  
ggplot(daily_data, aes(x = ARREST_DATE, y = Count)) +
  geom_line(size = 0.8, alpha = 0.8) +
  labs(title = "Daily NYPD Arrests",
       subtitle = "From January 1st to September 30th",
       x = "Date", y = "Number of Arrests", color = "Borough") +
  scale_x_date(date_breaks = "2 week", date_labels = "%b %d") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        plot.subtitle = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))
```

2. Overall demographic distribution of arrests:

```{r}
age_count <- data |>
  group_by(AGE_GROUP) |>
  summarize(Count = n())

plot1 <- ggplot(age_count, aes(x = AGE_GROUP, y = Count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  ggtitle("Number of Arrests by Age group") +
  xlab("Perp Age Group") +
  ylab("Number of Arrests")

gender_counts <- data |>
  group_by(PERP_SEX) |>
  summarize(Count = n()) |>
  mutate(Percentage = Count / sum(Count) * 100,
         Label = paste0(PERP_SEX, " ", round(Percentage, 1), "%"),
         Position = cumsum(Count) - Count / 2)

plot2 <- ggplot(gender_counts, aes(x = 2, y = Count, fill = PERP_SEX)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  labs(title = "Perp Gender") +
  theme_void() +
  scale_fill_manual(values = c("steelblue", "lightblue")) +
  geom_text(aes(x = 3, y = Position, label = Label), color = "black", size = 5) + 
  theme(legend.position = "none") + xlim(0.5, 3)

race_count <- data |>
  group_by(PERP_RACE) |>
  summarize(Count = n()) |>
  arrange(desc(Count))

plot3 <- ggplot(race_count, aes(x = reorder(PERP_RACE, -Count), y = Count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  ggtitle("Number of Arrests by PERP_RACE") +
  xlab("PERP_RACE") +
  ylab("Number of Arrests") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

top_15_ofns_desc <- data |>
  group_by(OFNS_DESC) |>
  summarise(Count = n()) |>
  arrange(desc(Count)) |>
  head(15)

plot4 <- ggplot(top_15_ofns_desc, aes(x = reorder(OFNS_DESC, Count), 
                                      y = Count)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(title = "Top 15 Arrest Categories",
       x = "Offense Description", y = "Number of Arrests") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  coord_flip()
```

```{r}
layout <- "\nABD\nCCD"

(plot1 + plot2 + plot3) + plot4 + plot_layout(design = layout, widths = c(1, 1))
```

This plot provides a comprehensive overview of demographic patterns and crime-related data. The top-left plot displays the distribution of arrests by age group, revealing that individuals aged `25-44` represent the highest proportion of arrests, followed by the `18-24` and `45-64` age groups. The pie chart at the top-right illustrates the gender distribution of arrests, with males accounting for a dominant 82.1% of all arrests, compared to 17.9% for females. The bottom-left plot focuses on the perpetrator’s race description, showing that `Black` individuals are most frequently arrested, followed by `White Hispanics`, `Black Hispanics`, and other racial categories. Finally, the bar chart on the bottom-right lists the top 15 offense categories, where `Assault 3 & Related Offenses` leads, followed by `Petit Larceny` and `Felony Assault`.

```{r}
alluvia_data <- na.omit(data) |> 
  filter(!(PERP_RACE %in% c("UNKNOWN", "AMERICAN INDIAN/ALASKAN NATIVE"))) |>
  group_by(PERP_SEX, PERP_RACE, AGE_GROUP) |>
  summarise(Count = n(), .groups = "drop") |>
  rename(Sex = PERP_SEX, Race = PERP_RACE, AgeGroup = AGE_GROUP)

ggplot(alluvia_data, aes(axis1 = Sex, axis2 = Race, 
                         axis3 = AgeGroup, y = Count)) +
  geom_alluvium(aes(fill = Race), width = 0.2, alpha = 0.8) +
  geom_stratum(width = 0.2, fill = "lightgrey", color = "black") +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2) +
  scale_x_discrete(limits = c("Sex", "Race", "AgeGroup"), expand = c(0.1, 0.1)) +
  theme_minimal() +
  scale_fill_brewer(type='qual') + 
  labs(title = "Flow of Arrests by Demographics",
       x = "Demographic Categories",
       y = "Number of Arrests", fill = "Race")
```

Among Black females, the majority of arrests are concentrated in the 25-44 age group, with a smaller proportion in the 18-24 and 45-64 age groups. Arrests of Black females under 18 or over 65 are minimal.

```{r}
# data1 <- data |> filter(!is.na(LAW_CAT_CD)) |> 
#   filter(!(PERP_RACE %in% c("UNKNOWN", "AMERICAN INDIAN/ALASKAN NATIVE"))) |>
#   filter(!LAW_CAT_CD %in% c("Violation")) |>
#   mutate(ARREST_BORO = as.factor(ARREST_BORO),
#          LAW_CAT_CD = as.factor(LAW_CAT_CD),
#          PERP_RACE = as.factor(PERP_RACE))

mosaic_data <- data |> filter(!is.na(LAW_CAT_CD)) |> 
  filter(!(PERP_RACE %in% c("UNKNOWN", "AMERICAN INDIAN/ALASKAN NATIVE")))

fill_colors <- RColorBrewer::brewer.pal(3, "Blues")

# Generate the mosaic plot
mosaic(LAW_CAT_CD ~ ARREST_BORO + PERP_RACE, data = mosaic_data,
       labeling_args = list(
         set_varnames = c(
           ARREST_BORO = "Borough", 
           PERP_RACE = "Race", 
           LAW_CAT_CD = "Offense Level"
         )),
       direction = c("v","h","h"),
       highlighting_fill = fill_colors,
       main = "Mosaic Plot: Borough, Race, and Offense Level")
```

=======
# Temporal Analysis of Arrests: Time Series Trends
```{r}
data <- data |>
  mutate(Weekday = wday(ARREST_DATE, label = TRUE, abbr = TRUE))  

weekday_summary <- data |>
  group_by(ARREST_DATE, Weekday) |>
  summarise(Count = n(), .groups = "drop")  

ggplot(weekday_summary, aes(x = ARREST_DATE, y = Count, group = Weekday, fill = Count)) +
  geom_area(alpha = 0.8) +  
  geom_line(size = 1, color = "black") +  
  facet_grid(Weekday ~ ., scales = "fixed", switch = "y") +  
  scale_fill_gradient(
    low = "lightblue",  
    high = "darkblue",  
    name = "Arrest Count"  
  ) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    title = "Daily NYPD Arrests by Weekday",
    x = "Date",
    y = "Number of Arrests"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1), 
    strip.text.y.left = element_text(angle = 0),      
    strip.placement = "outside",                     
    legend.position = "right"                    
  )
```


```{r}
daily_summary_by_borough <- data |>
  group_by(ARREST_DATE, ARREST_BORO) |>
  summarise(Count = n(), .groups = "drop")

wide_data <- tidyr::pivot_wider(daily_summary_by_borough, 
                                names_from = ARREST_BORO, 
                                values_from = Count, 
                                values_fill = 0)

wide_data$ARREST_DATE <- as.Date(wide_data$ARREST_DATE)

time_series_obj <- xts(wide_data[,-1], order.by = wide_data$ARREST_DATE)

custom_colors <- c(
  "Bronx" = "#E7298A",    
  "Brooklyn" = "#6495ED", 
  "Manhattan" = "#E6AB02",
  "Queens" = "#66A61E",   
  "Staten Island" = "#7570B3" 
)

dygraph(time_series_obj, main = "Daily NYPD Arrests") |>
  dyAxis("y", label = "Number of Arrests", valueRange = c(0, 400)) |>
  dyAxis("x", label = "Date") |>
  dyRangeSelector() |>
  dyLegend(width = 300, labelsSeparateLines = TRUE) |>
  dyOptions(colors = unname(custom_colors), strokeWidth = 2, gridLineColor = "#DDDDDD") 

```

```{r fig.width = 12}
daily_summary_by_borough <- data |>
  group_by(ARREST_DATE, ARREST_BORO) |>
  summarise(Count = n(), .groups = "drop")

ggplot(daily_summary_by_borough, aes(x = ARREST_DATE, y = Count, color = ARREST_BORO)) +
  geom_line(size = 0.8, alpha = 0.8) +
  labs(
    title = "Daily NYPD Arrests",
    subtitle = "From January 1st to September 30th",
    x = "Date",
    y = "Number of Arrests",
    color = "Borough"
  ) +
  scale_x_date(date_breaks = "2 week", date_labels = "%b %d") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

```{r fig.width = 12}
weekly_summary_by_borough <- data |>
  mutate(Week = floor_date(ARREST_DATE, unit = "week")) |>  
  group_by(Week, ARREST_BORO) |>                           
  summarise(Count = n(), .groups = "drop") 
max_week <- max(weekly_summary_by_borough$Week)
weekly_summary_filtered <- weekly_summary_by_borough |>
  filter(Week < max_week)

ggplot(weekly_summary_filtered, aes(x = Week, y = Count, color = ARREST_BORO)) +
  geom_line(size = 1, alpha = 0.8) +
  scale_x_date(date_breaks = "2 week", date_labels = "%b %d") + 
  labs(
    title = "Weekly NYPD Arrests",
    subtitle = "From January 1st to September 30th",
    x = "Week",
    y = "Number of Arrests",
    color = "Borough"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1), 
    axis.title.x = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```


